### 2025-03-14 paper/quarto_ecml/paper.tex
Parse key comments from second/third document
TODO Patrick: remaining comments, namely (1) better labels for Fig. 1? (2) reorganizing Table 1? (3) potentially some extra explanation in results?

This interpretation of the link between explainability through
counterfactuals on one side and robustness to adversarial examples on
the other is backed by empirical evidence.
\cite{sauer2021counterfactual} demonstrate that using counterfactual
images during classifier training improves model robustness. Similarly,
\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
potentially useful training data in machine learning, especially in
supervised settings where inputs may be reasonably mapped to multiple
outputs. They, too, demonstrate that augmenting the training data of
image classifiers can improve generalization. Finally,
\cite{teney2020learning} propose an approach using counterfactuals in
training that does not rely on data augmentation: they argue that
counterfactual pairs typically already exist in training datasets.
Specifically, their approach relies on identifying similar input samples
with different annotations and ensuring that the gradient of the
classifier aligns with the vector between such pairs of counterfactual
inputs using the cosine distance as the loss function.

@@

As observed by \cite{grathwohl2020your}, any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first term can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimize the contrastive divergence between
these observed samples from \(p(\mathbf{x})\) and generated samples from
\(p_\theta(\mathbf{x})\).

A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in Def.
\ref{def-explainability}. This follows directly if we consider samples
drawn from \(p_\theta(\mathbf{x})\) as counterfactuals because the JEM
objective effectively minimizes the divergence between the conditional
posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
\cite{grathwohl2020your} rely on Stochastic Gradient Langevin Dynamics
(SGLD) using an uninformative prior for initialization but we depart
from their methodology. Instead of SGLD, we propose to use
counterfactual explainers to generate counterfactuals of observed
training samples. Specifically, we have:

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
function. We set
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
where \(y^+\) denotes the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
sampled factual from a non-target class generated with a gradient-based
CE generator for up to \(T\) iterations. Mature counterfactuals are ones
that have either reached convergence wrt. the decision threshold
\(\tau\) or exhausted \(T\).

@@

The departure from SGLD allows us to tap into the vast repertoire of
explainers that have been proposed in the literature to meet different
desiderata. For example, many methods facilitate the imposition of
domain and mutability constraints. In principle, any existing approach
for generating counterfactual explanations is viable, so long as it does
not violate the faithfulness condition. Like JEMs
\cite{murphy2022probabilistic}, CT can be considered a form of
contrastive representation learning.

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
\cite{luu2023counterfactual,balashankar2023improving} or AEs
\cite{freiesleben2022intriguing,pawelczyk2022exploring}. This leaves
some flexibility with respect to the choice for
\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
functional form, but likely not the only sensible choice, is inspired by
adversarial training:

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \cite{szegedy2013intriguing} who define an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we choose to work with a different distinction between CE and AE
than \cite{freiesleben2022intriguing} who consider misclassification as
the key distinguishing feature of AE. One of the key observations in
this work is that we can leverage CEs during training and get
adversarial examples essentially for free.

@@

We have simulated this scenario using synthetic data with two
independent features and Gaussian class-conditional densities in
Figure~\ref{fig-poc}. The four panels show the outcomes for different
training procedures using the same model architectures (a linear
classifier). In panels (a) and (c) we have trained the models
conventionally, while in panels (b) and (d) we used CT. Only in panels
(c) and (d) do we impose the mutability constraint on \emph{age} at test
time. In each case, we show the decision boundary (in green) and the
training data colored according to their ground-truth label: orange
points belong to the target class, \(y^+=2\), blue points belong to the
non-target class, \(y^-=1\). Stars indicate CEs in the target class
generated at test time using generic gradient descent until convergence.

@@

Here we present the experiments conducted to answer our research
questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does our counterfactual training objective as defined
  in Equation~\ref{eq-obj} induce models to learn plausible
  explanations?
\item
  To what extent does the CT objective produce more favorable
  algorithmic recourse outcomes in the presence of actionability
  constraints?
\item
  What are the effects of hyperparameter selection wrt. the CT
  objective?
\end{enumerate}

@@

Our key outcome of interest is how well do models perform with respect
to explainability (Def. \ref{def-explainability}). To this end, we focus
primarily on the plausibility and cost of faithfully generated
counterfactuals at test time. To measure the cost of counterfactuals, we
follow the standard convention of using distances (\(\ell_1\)-norm)
between factuals and counterfactuals as a proxy. For plausibility, we
assess how similar counterfactuals are to observed samples in the target
domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
distance-based metric used by \cite{altmeyer2024faithful},

@@

In Section~\ref{sec-method}, we show that our proposed way for encoding
mutability constraints leads to lower classifier sensitivity wrt.
immutable features for linear models, tilting the decision boundary in
favour of mutable features instead. For binding constraints at test
time, this has the effect of shorter counterfactual paths and hence
smaller average costs to individuals. To extend this to the non-linear
case, we test the effect of imposing mutability constraints empirically
for our synthetic data using the same evaluation scheme as above. The
final row in Table~\ref{tbl-main} reports the average reduction in costs
for CT compared to the baseline, when imposing that either the first or
the second feature is immutable. In all cases, costs are reduced
substantially indicating that classifiers trained with CT are indeed
more sensitive to mutable features.

>>>>

This interpretation of the link between explainability through
counterfactuals on one side and robustness to adversarial examples on
the other is backed by empirical evidence.
\cite{sauer2021counterfactual} demonstrates that using counterfactual
images during classifier training improves model robustness. Similarly,
\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
potentially useful training data in machine learning, especially in
supervised settings where inputs may be reasonably mapped to multiple
outputs. They, too, demonstrate that augmenting the training data of
image classifiers can improve generalization. Finally,
\cite{teney2020learning} propose an approach using counterfactuals in
training that does not rely on data augmentation: they argue that
counterfactual pairs typically already exist in training datasets.
Specifically, their approach relies on identifying similar input samples
with different annotations and ensuring that the gradient of the
classifier aligns with the vector between such pairs of counterfactual
inputs using the cosine distance as the loss function.

@@

As observed by \cite{grathwohl2020your}, any classifier can be
re-interpreted as a joint energy-based model (JEM) that learns to
discriminate output classes conditional on the observed (training)
samples from \(p(\mathbf{x})\) and the generated samples from
\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
perform well at both tasks by directly maximizing the joint
log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first term can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
\cite{grathwohl2020your} minimizes the contrastive divergence between
these observed samples from \(p(\mathbf{x})\) and generated samples from
\(p_\theta(\mathbf{x})\).

A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
tend to do well with respect to the plausibility objective in Def.
\ref{def-explainability}. This follows directly if we consider samples
drawn from \(p_\theta(\mathbf{x})\) as counterfactuals because the JEM
objective effectively minimizes the divergence between the conditional
posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
\cite{grathwohl2020your} relies on Stochastic Gradient Langevin Dynamics
(SGLD) using an uninformative prior for initialization but we depart
from their methodology. Instead of SGLD, we propose to use
counterfactual explainers to generate counterfactuals of observed
training samples. Specifically, we have:

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
function defined as
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
where \(y^+\) denotes the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
sampled factual from a non-target class generated with a gradient-based
CE generator for up to \(T\) iterations. Mature counterfactuals are ones
that have either reached convergence wrt. the decision threshold
\(\tau\) or exhausted \(T\).

@@

The departure from SGLD of \cite{grathwohl2020your} allows us to tap
into the vast repertoire of explainers that have been proposed in the
literature to meet different desiderata. For example, many methods
facilitate the imposition of domain and mutability constraints. In
principle, any existing approach for generating counterfactual
explanations is viable, so long as it does not violate the faithfulness
condition. Like JEMs \cite{murphy2022probabilistic}, CT can be
considered a form of contrastive representation learning.

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
\cite{luu2023counterfactual,balashankar2023improving} or adversarial
examples \cite{freiesleben2022intriguing,pawelczyk2022exploring}. This
leaves some flexibility wrt. the choice for \(\text{advloss}(\cdot)\) in
Equation~\ref{eq-obj}. An intuitive functional form, but likely not the
only sensible choice, is inspired by adversarial training:

\begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with \cite{szegedy2013intriguing} who define an
adversarial attack as an ``imperceptible non-random perturbation''.
Thus, we choose to work with a different distinction between CE and AE
than \cite{freiesleben2022intriguing} who consider misclassification as
the key distinguishing feature of AE. One of the key observations of
this work is that we can leverage CEs during training and get
adversarial examples essentially for free, which can be used to reap the
aforementioned benefits of adversarial training.

@@

We have simulated this scenario using synthetic data with independent
\emph{age} and \emph{debt}, and Gaussian class-conditional densities in
Figure~\ref{fig-poc}. The four panels show the outcomes for different
training procedures using the same model architectures (a linear
classifier). In panels (a) and (c) we have trained the models
conventionally, while in panels (b) and (d) we used CT. Only in panels
(c) and (d) do we impose the mutability constraint on \emph{age} at test
time. In each case, we show the decision boundary (in green) and the
training data colored according to their ground-truth label: orange
points belong to the target class, \(y^+=2\), blue points belong to the
non-target class, \(y^-=1\). Stars indicate CEs in the target class
generated at test time using generic gradient descent until convergence.

@@

In our experiments we seek to answer the following three research
questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  To what extent does the counterfactual training objective (as defined
  in Equation~\ref{eq-obj}) induce models to learn plausible
  explanations?
\item
  To what extent does the CT objective produce more favorable
  algorithmic recourse outcomes in the presence of actionability
  constraints?
\item
  What are the effects of hyperparameter selection wrt. the CT
  objective?
\end{enumerate}

@@

Our key outcome of interest is how well models perform with respect to
explainability (Def. \ref{def-explainability}). To this end, we focus
primarily on the plausibility and cost of faithfully generated
counterfactuals at test time. To measure the cost of counterfactuals, we
follow the standard convention of using distances (\(\ell_1\)-norm)
between factuals and counterfactuals as a proxy. For plausibility, we
assess how similar counterfactuals are to observed samples in the target
domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
distance-based metric used by \cite{altmeyer2024faithful},

@@

In Section~\ref{sec-method}, we show that our proposed way for encoding
mutability constraints leads to lower classifier sensitivity wrt.
immutable features for linear models, tilting the decision boundary in
favour of mutable features instead. For binding constraints at test
time, this has the effect of shorter counterfactual paths and hence
smaller average costs (\(\ell_1\)-norm) to individuals. To extend this
to the non-linear case, we test the effect of imposing mutability
constraints empirically for our synthetic data using the same evaluation
scheme as above. The final row in Table~\ref{tbl-main} reports the
average reduction in costs for CT compared to the baseline, when
imposing that either the first or the second feature is immutable. In
all cases, costs are reduced substantially, indicating that classifiers
trained with CT are indeed more sensitive to mutable features.

### 2025-03-14 paper/sections/experiments.qmd
Parse key comments from second/third document
TODO Patrick: remaining comments, namely (1) better labels for Fig. 1? (2) reorganizing Table 1? (3) potentially some extra explanation in results?

Here we present the experiments conducted to answer our research questions:

1. To what extent does our counterfactual training objective as defined in @eq-obj induce models to learn plausible explanations?
2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection wrt. the CT objective?

@@

Our key outcome of interest is how well do models perform with respect to explainability (Def. \ref{def-explainability}). To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar counterfactuals are to observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the distance-based metric used by @altmeyer2024faithful,

@@

In @sec-method, we show that our proposed way for encoding mutability constraints leads to lower classifier sensitivity wrt. immutable features for linear models, tilting the decision boundary in favour of mutable features instead. For binding constraints at test time, this has the effect of shorter counterfactual paths and hence smaller average costs to individuals. To extend this to the non-linear case, we test the effect of imposing mutability constraints empirically for our synthetic data using the same evaluation scheme as above. The final row in @tbl-main reports the average reduction in costs for CT compared to the baseline, when imposing that either the first or the second feature is immutable. In all cases, costs are reduced substantially indicating that classifiers trained with CT are indeed more sensitive to mutable features.

>>>>

In our experiments we seek to answer the following three research questions:

1. To what extent does the counterfactual training objective (as defined in @eq-obj) induce models to learn plausible explanations?
2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection wrt. the CT objective?

@@

Our key outcome of interest is how well models perform with respect to explainability (Def. \ref{def-explainability}). To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar counterfactuals are to observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the distance-based metric used by @altmeyer2024faithful,

@@

In @sec-method, we show that our proposed way for encoding mutability constraints leads to lower classifier sensitivity wrt. immutable features for linear models, tilting the decision boundary in favour of mutable features instead. For binding constraints at test time, this has the effect of shorter counterfactual paths and hence smaller average costs ($\ell_1$-norm) to individuals. To extend this to the non-linear case, we test the effect of imposing mutability constraints empirically for our synthetic data using the same evaluation scheme as above. The final row in @tbl-main reports the average reduction in costs for CT compared to the baseline, when imposing that either the first or the second feature is immutable. In all cases, costs are reduced substantially, indicating that classifiers trained with CT are indeed more sensitive to mutable features.

### 2025-03-14 paper/sections/method.qmd
Parse key comments from second/third document
TODO Patrick: remaining comments, namely (1) better labels for Fig. 1? (2) reorganizing Table 1? (3) potentially some extra explanation in results?

As observed by @grathwohl2020your, any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$. The first term can be optimized using conventional cross-entropy as in @eq-obj. Then, to optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between these observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.

A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in Def. \ref{def-explainability}. This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization but we depart from their methodology. Instead of SGLD, we propose to use counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:

$$
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
$$ {#eq-div}
where $\mathcal{E}_\theta(\cdot)$ denotes the energy function. We set $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$. Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations. Mature counterfactuals are ones that have either reached convergence wrt. the decision threshold $\tau$ or exhausted $T$.

@@

The departure from SGLD allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata. For example, many methods facilitate the imposition of domain and mutability constraints. In principle, any existing approach for generating counterfactual explanations is viable, so long as it does not violate the faithfulness condition. Like JEMs [@murphy2022probabilistic], CT can be considered a form of contrastive representation learning.

@@

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual;@balashankar2023improving] or AEs [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility with respect to the choice for $\text{advloss}(\cdot)$ in @eq-obj. An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:

$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}
Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing who consider misclassification as the key distinguishing feature of AE. One of the key observations in this work is that we can leverage CEs during training and get adversarial examples essentially for free.

@@

We have simulated this scenario using synthetic data with two independent features and Gaussian class-conditional densities in @fig-poc. The four panels show the outcomes for different training procedures using the same model architectures (a linear classifier). In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we used CT. Only in panels (c) and (d) do we impose the mutability constraint on *age* at test time. In each case, we show the decision boundary (in green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$. Stars indicate CEs in the target class generated at test time using generic gradient descent until convergence.

>>>>

As observed by @grathwohl2020your, any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$. The first term can be optimized using conventional cross-entropy as in @eq-obj. Then, to optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimizes the contrastive divergence between these observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.

A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in Def. \ref{def-explainability}. This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your relies on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization but we depart from their methodology. Instead of SGLD, we propose to use counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:

$$
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
$$ {#eq-div}
where $\mathcal{E}_\theta(\cdot)$ denotes the energy function defined as $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$. Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations. Mature counterfactuals are ones that have either reached convergence wrt. the decision threshold $\tau$ or exhausted $T$.

@@

The departure from SGLD of [@grathwohl2020your] allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata. For example, many methods facilitate the imposition of domain and mutability constraints. In principle, any existing approach for generating counterfactual explanations is viable, so long as it does not violate the faithfulness condition. Like JEMs [@murphy2022probabilistic], CT can be considered a form of contrastive representation learning.

@@

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual;@balashankar2023improving] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility wrt. the choice for $\text{advloss}(\cdot)$ in @eq-obj. An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:

$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}
Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing who consider misclassification as the key distinguishing feature of AE. One of the key observations of this work is that we can leverage CEs during training and get adversarial examples essentially for free, which can be used to reap the aforementioned benefits of adversarial training.

@@

We have simulated this scenario using synthetic data with independent *age* and *debt*, and Gaussian class-conditional densities in @fig-poc. The four panels show the outcomes for different training procedures using the same model architectures (a linear classifier). In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we used CT. Only in panels (c) and (d) do we impose the mutability constraint on *age* at test time. In each case, we show the decision boundary (in green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$. Stars indicate CEs in the target class generated at test time using generic gradient descent until convergence.
